---
title: "Data Processing for Low Carb High Fat Diet"
subtitle: "Repeated measures on subjects"
author: "Mark Connolly"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html:
    toc: true
    toc_depth: 2
    citation_package: natbib
    latex_engine: xelatex
    highlight: tango
bibliography: [lchf.bib,zotero.bib]
link-citations: yes
---

```{r setup, include=FALSE}
library(knitr)
library(tufte)
library(rmarkdown)

library(ggplot2)
library(readxl)
library(gdata)
library(tidyr)
library(dplyr)
library(magrittr)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```
# Introduction
Analysis of collected observations.  Weaving executed code with explanatory text to produce literate programming [@knuth_literate_1984] is provided by the R packages `knitr` [@R-knitr] and the document template `tufte`  [@R-tufte].

This document is translated from rmarkdown [@R-rmarkdown] source.  The source can be found at [LCHF Analysis.Rmd](https://github.com/markconnolly/lchf/blob/master/LCHF%20Analysis.Rmd).

The `%>%` operator will be seen in some of the code.  The package `magrittr` [@R-magrittr] provides this operator, called a pipe, which forwards the result of a preceeding function to a suceeding function in a sinle code stream.  By default, the forwarded object is "piped" into the succeeding function's first parameter slot.  The argument can be moved to other parameter slots in the succeeding call Using the `.` placeholder.  The pipe is used where it improves readability.  A series of functions interspersed with `%>%` should be viewed as pipeline or processing elements, with the output of one function the input to the next.

R uses formulae to define response models.  A formula is the form 

```response variable ~ explanatory variables```

The structure of formulae is well-explained by [@king_r_formulae].

# Crossover
At the start of the test, fat will predominate as the source of calories.  As the test proceeds and the energy burn rate increases (as indicated by increased heartrate), fat as a source will diminish and carbohydrate as a source will increase.  The crossover graph captures the trends and imputes the heartrate where fat and carbohydrate contribution are equal (the point where carbohydrate will start to become the predominant energy source).

## Data preparation for crossover
Set up for combining several spreadsheets into a single data frame.  The sheets must all have the same naming convention, the same sheet structure, and the same column names.

```{r}
# a filter to be used to tell the Windows operating system to limit 
# file selection to files that look like Excel
excelfilter <- matrix(c("Excel", "*.xls*"), ncol=2)

# a function that is used to ingest a conventionally named 
# and structured spreadsheet and convert the file to a 
# data frame (which is returned to the caller)
ingestvo2 <- function(afilename) {
  nameparts <- afilename %>% basename() %>% strsplit("[.]") %>% unlist()
  columnsfromfilename <- 
    data.frame("subj" = as.character(nameparts[1]),
               "testnumber" = as.character(nameparts[2]),
               "testdate" = as.Date(nameparts[3],format="%m%d%y")
              ) 
  sheet <- 
    read_excel(afilename, sheet=1, skip=4, col_names=FALSE)
  
  names(sheet) <- 
    trim(names(read_excel(afilename, 
                          sheet=1, 
                          skip=2, 
                          col_names=TRUE)))

  return(cbind(columnsfromfilename,sheet))
}
```

## Select and combine Excel source files
Create a list that contains user-selected files, with each file processed by the `ingestvo2` function.  The data frames in the list are then bound together by row.  This is achieved in code using a pipeline of functions to process the inputs before assigning to the target variable. Note the `%>%` operator.

```{r}
vo2data <- 
  choose.files(filters = excelfilter,
               caption = "Select VO2 files to combine") %>%
  lapply(FUN=ingestvo2) %>% 
  do.call('rbind', .)

vo2data$subj = as.factor(vo2data$subj)
```


## Subset the test subject and test number

```{r}
fuel <- 
  vo2data[,c("testnumber","subj","%Fat","%Carbs","HR")]
knitr::kable(head(fuel))
```

## Restructure the data frame
Using the `gather` [@R-tidyr] function, create a "long" form of the data to be used for the plot.  The `gather` function combines two or more columns such that the column names become factor values in a new column (the key colum) and the values of the designated columns become values in a second new column (the value column). The columns from the source not involved in the gathering remain in place, with values duplicated as needed to fill out the long structure. Functions `head` is used to convey the structure changes to the data frame, which can be compared to `head(fuel)` from above.  The new structure makes "%Fat" and "%Carbs" factors and their values collapsed into a single variable.  

```{r warning=FALSE}
fuel_long <- gather(fuel,                # source data frame
                    percentage,          # new column name of the key
                    pctKcalmin,          # new column name of the value
                    c(`%Fat`,`%Carbs`))  # columns that are sources of
                                         # key names and values.
knitr::kable(head(fuel_long))   # shows the top part of the data frame
```

## Predicted calorie source ratio by heartrate
The percent contribution of fat and carbohydrates are plotted against heartrate.  Plots are faceted by test number and subject to show individual responses across tests.  The plots are created using `ggplot2` [@R-ggplot2].  

```{r}
ggplot(data=fuel_long, 
       aes(x=HR, y=pctKcalmin, 
           color=percentage, 
           shape=percentage)) + 
  geom_point() + 
  geom_smooth(method="loess") +
  facet_grid(subj ~ testnumber)
```

##  Predicted heartrate based on calorie source with predicted crossover
Using data manipulation functions from `dplyr` [@R-dplyr], the observations are used to build prediction models for each calorie source for each subject for each test.  The prediction models are used to predict the heartrate when calorie sources are equal (both fat and carbohydrates are at 50%).  The resulting data frame is used to mark and label the plots at the predicted heartrate. 

```{r}
crossoverHR50 <- group_by(fuel_long, testnumber, subj, percentage) %>% 
                 do(HR = round(predict(loess(HR ~ pctKcalmin, data = .), newdata=50))) %>%
                 unnest() %>% 
                 as.data.frame()

knitr::kable(crossoverHR50, 
             caption = "Predicted heartrate at crossover")

ggplot(data=fuel_long,
       aes(x=pctKcalmin,
           y=HR,
           color=percentage, 
           shape=percentage)) + 
  geom_point() + 
  geom_smooth() +
  facet_grid(subj ~ testnumber) +
  geom_hline(data = crossoverHR50, aes(yintercept = HR), colour="darkgrey") + 
  geom_text(data = crossoverHR50, aes(label=HR, x=0, y=HR), colour="black") +
  coord_flip()
```

```{r include=FALSE}
# References
citPkgs <- c(names(sessionInfo()$otherPkgs), "base")
write_bib(citPkgs, file="lchf.bib")

```
# Heartrate at RER 0.5
Heartrate at RER 0.5 is the dependent variable.  The hypothesis is that the low carbohydrate/high fat diet will increase the heartrate at RER, with the body becoming using more fat than protien as the testing procedes.
```{r}
rer <- choose.files(filters = excelfilter,
                    caption = "Select rer file",
                    multi=FALSE) %>%
       read_excel(sheet=1)
# make the names more convenient by squeezing out spaces
names(rer) <- gsub(" ", "", names(rer), fixed=TRUE)

boxplot(rer$RERHR ~ rer$Subject)

rer$Subject <- factor(rer$Subject)
rer$TestNumber <- factor(rer$TestNumber)

summary(aov(RERHR ~ Error(Subject), data=rer))


```
