---
title: "Data Processing for Low Carb High Fat Diet"
subtitle: "Repeated measures on subjects"
author: "Mark Connolly"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html:
    toc: true
    toc_depth: 2
    citation_package: natbib
    latex_engine: xelatex
    highlight: tango
bibliography: [lchf.bib,zotero.bib]
link-citations: yes
---

```{r setup, include=FALSE}
library(knitr)
library(tufte)
library(rmarkdown)

library(ggplot2)
library(readxl)
library(gdata)
library(tidyr)
library(dplyr)
library(magrittr)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```
# Introduction
Analysis of collected observations.  Weaving executed code with explanatory text to produce literate programming [@knuth_literate_1984] is provided by the R packages `knitr` [@R-knitr] and the document template `tufte`  [@R-tufte].

This document is translated from rmarkdown [@R-rmarkdown] source.  The source can be found at [LCHF Analysis.Rmd](https://github.com/markconnolly/lchf/blob/master/LCHF%20Analysis.Rmd).

The `%>%` operator will be seen in some of the code.  The package `magrittr` [@R-magrittr] provides this operator, called a pipe, which forwards the result of a preceeding function to a suceeding function in a sinle code stream.  By default, the forwarded object is "piped" into the succeeding function's first parameter slot.  The argument can be moved to other parameter slots in the succeeding call Using the `.` placeholder.  The pipe is used where it improves readability.  A series of functions interspersed with `%>%` should be viewed as pipeline or processing elements, with the output of one function the input to the next.

R uses formulae to define response models.  A formula is the form 

```response variable ~ explanatory variables```

The structure of formulae is well-explained by [@king_r_formulae].

# Plots of observation variables

```{r}
excelfilter <- matrix(c("Excel", "*.xls*"), ncol=2)

rer <- choose.files(filters = excelfilter,
                    caption = "Select rer file",
                    multi=FALSE) %>%
       read_excel(sheet=1)

# make the names more convenient to use
names(rer) <- gsub(" ", "", names(rer), fixed=TRUE)
names(rer) <- gsub("(", "_", names(rer), fixed=TRUE)
names(rer) <- gsub(")", "", names(rer), fixed=TRUE)
names(rer) <- gsub("%", "pct", names(rer), fixed=TRUE)

# convert factors to factors
rer$Subject <- as.factor(rer$Subject)
rer$TestNumber <- as.factor(rer$TestNumber)

rer2 <- rer[c("Subject","TestNumber",
              "RERHR",
              "MaxRelVO2","MaxAbsVO2",
              "BFpctBIA","SFBFpct",
              "MoodB4_cm","MoodAfter_cm",
              "FatBIA","FatSF")]

rer2_long <- 
  rer[c("Subject","TestNumber",
        "RERHR",
        "MaxRelVO2","MaxAbsVO2",
        "BFpctBIA","SFBFpct",
        "MoodB4_cm","MoodAfter_cm",
        "FatBIA","FatSF")]            %>%
  gather(measurement,obs,RERHR:FatSF)

rer2_long$measurement <- 
  ordered(rer2_long$measurement, 
          levels = c("RERHR",
                     "MaxRelVO2","MaxAbsVO2",
                     "BFpctBIA","SFBFpct",
                     "MoodB4_cm","MoodAfter_cm",
                     "FatBIA","FatSF"))

scales <- "fixed";columns <- c("RERHR")
rer2_long %>% 
  subset(measurement %in% columns) %>% 
  ggplot(data=., aes(x=TestNumber,
                     y=obs,
                     color=Subject,
                     group = Subject)) + 
  geom_point() + 
  geom_line() +
  facet_wrap(~measurement, scales = scales)


scales <- "free";columns <- c("MaxRelVO2","MaxAbsVO2")
rer2_long %>% 
  subset(measurement %in% columns) %>% 
  ggplot(data=., aes(x=TestNumber,
                     y=obs,
                     color=Subject,
                     group = Subject)) + 
  geom_point() + 
  geom_line() +
  facet_wrap(~measurement, scales = scales)


scales <- "fixed";columns <- c("BFpctBIA","SFBFpct")
rer2_long %>% 
  subset(measurement %in% columns) %>% 
  ggplot(data=., aes(x=TestNumber,
                     y=obs,
                     color=Subject,
                     group = Subject)) + 
  geom_point() + 
  geom_line() +
  facet_wrap(~measurement, scales = scales)


scales <- "fixed";columns <- c("MoodB4_cm","MoodAfter_cm")
rer2_long %>% 
  subset(measurement %in% columns) %>% 
  ggplot(data=., aes(x=TestNumber,
                     y=obs,
                     color=Subject,
                     group = Subject)) + 
  geom_point() + 
  geom_line() +
  facet_wrap(~measurement, scales = scales)


scales <- "fixed";columns <- c("FatBIA","FatSF")
rer2_long %>% 
  subset(measurement %in% columns) %>% 
  ggplot(data=., aes(x=TestNumber,
                     y=obs,
                     color=Subject,
                     group = Subject)) + 
  geom_point() + 
  geom_line() +
  facet_wrap(~measurement, scales = scales)





```

# Heartrate at RER 0.5
For later

Heartrate at RER 0.5 is the dependent variable.  The hypothesis is that the low carbohydrate/high fat diet will increase the heartrate at RER, with the body becoming using more fat than protien as the testing procedes.

Working model for analysis: [@gribble_repeated_2009]

`summary(aov(RERHR ~ Error(Subject), data=rer))`


```{r include=FALSE}
# References
citPkgs <- c(names(sessionInfo()$otherPkgs), "base")
write_bib(citPkgs, file="lchf.bib")

```
