---
title: "Paired t-tests for LCHF study"
output: 
  html_notebook: 
    toc: TRUE
    theme: "cosmo"
---
# Repeated measures low carb high fat diet


## Packages
Load the various packages.

```{r setup}

# packages for analysis of repeated measures
library(nlme)
library(car)

# document definition and construction
library(knitr)
opts_knit$set(eval.after = "fig.cap")
library(rmarkdown)

#reading and writing files of various types (in this case, csv and text files)
library(readr)

# data manipulation
library(tidyr)
library(dplyr)
library(tibble)

# pipelining commands
library(magrittr)

# plots
library(ggplot2)

```

## Prepare data

```{r}
obs <- read_csv("mep_observations.csv")

# these models work with factors
obs$Subject <- as.factor(obs$Subject)
obs$TestNumber <- as.factor(obs$TestNumber)
obs <- as_tibble(obs)
obs$Week <- obs$TestNumber
levels(obs$Week) <- c("Baseline", "Week1", "Week2", "Week3")

obs <- obs[,c("Subject", "TestNumber", "Week",   # move Week next to TestNumber
              "MEPHR", "Weight", "MaxRelVO2",    # for easy validation
              "FatBIA", "FatSF", 
              "Avgkcals", "AvgFat", "AvgCHO", "AvgPro")]

obs
```





***
## Average KCal
### Linear mixed-effects model
```{r}
lme.model <- lme(Avgkcals ~ Week, 
                 random = ~1|Subject/Week, 
                 data=obs)
summary(lme.model)
```

```{r}
# print F- and p-values
anova(lme.model)
```


### Boxplot
```{r}
aboxplot <- ggplot(data=obs, aes(x=Week,y=Avgkcals)) + 
            geom_boxplot() +
            labs(x="", y=expression("Mean kcal/day"))

aboxplot
```

### Pairwise t-test
```{r}
pairwise.t.test(x = obs$Avgkcals,
                g = obs$Week,
                p.adjust.method = "bonferroni",
                pool.sd = FALSE,
                paired = TRUE )
```
### Multivariate approach using lm() / mlm()
1. Convert data frame to matrix. Each row of the matrix is a subject.  Each column is the test number.
```{r}
obs.matrix <- with(obs, # data from obs data frame
                    cbind(
                      Avgkcals[Week=="Baseline"], # subject observations baseline
                      Avgkcals[Week=="Week1"], # subject observations test 1
                      Avgkcals[Week=="Week2"], # etc
                      Avgkcals[Week=="Week3"]
                      ) # cbind column bind, rows are flipped to columns
                  )
# set row and column identities
rownames(obs.matrix) <- levels(obs$Subject)
colnames(obs.matrix) <- levels(obs$Week)

# display the matrix
obs.matrix
```
2. Create a multivariate linear model from the matrix.
```{r}
mlm.model <- lm(obs.matrix ~ 1)
mlm.model

```

The coefficients are equal to the means across subjects at each test.

3. Create a set of factors.
```{r}
rfactor <- as.factor(levels(obs$Week))
rfactor
```

4. Define a new anova model object ```mlm.model.aov```.
```{r}
mlm.model.aov <- Anova(mlm.model, # the multivariate linear model
                       idata = data.frame(rfactor),
                       idesign = ~rfactor, 
                       type = "III")

summary(mlm.model.aov, 
        multivariate = FALSE) # don't show multivariate tests
```







***
## Average Fat
### Linear mixed-effects model
```{r}
lme.model <- lme(AvgFat ~ Week, 
                 random = ~1|Subject/Week, 
                 data=obs)
summary(lme.model)
```

```{r}
# print F- and p-values
anova(lme.model)
```

### Boxplot
```{r}
aboxplot <- ggplot(data=obs, aes(x=Week,y=AvgFat)) + 
            geom_boxplot() +
            annotate(geom="text", label="*", x=2.01, y=180, size=6, family="serif") +
            labs(x="", y=expression("Mean Fat g/day"))

aboxplot
```

### Pairwise t-test
```{r}
pairwise.t.test(x = obs$AvgFat,
                g = obs$Week,
                p.adjust.method = "bonferroni",
                pool.sd = FALSE,
                paired = TRUE )
```
### Multivariate approach using lm() / mlm()
1. Convert data frame to matrix. Each row of the matrix is a subject.  Each column is the test number.
```{r}
obs.matrix <- with(obs, # data from obs data frame
                    cbind(
                      AvgFat[Week=="Baseline"], # subject observations baseline
                      AvgFat[Week=="Week1"], # subject observations test 1
                      AvgFat[Week=="Week2"], # etc
                      AvgFat[Week=="Week3"]
                      ) # cbind column bind, rows are flipped to columns
                  )
# set row and column identities
rownames(obs.matrix) <- levels(obs$Subject)
colnames(obs.matrix) <- levels(obs$Week)

# display the matrix
obs.matrix
```
2. Create a multivariate linear model from the matrix.
```{r}
mlm.model <- lm(obs.matrix ~ 1)
mlm.model

```

The coefficients are equal to the means across subjects at each test.

3. Create a set of factors.
```{r}
rfactor <- as.factor(levels(obs$Week))
rfactor
```

4. Define a new anova model object ```mlm.model.aov```.
```{r}
mlm.model.aov <- Anova(mlm.model, # the multivariate linear model
                       idata = data.frame(rfactor),
                       idesign = ~rfactor, 
                       type = "III")

summary(mlm.model.aov, 
        multivariate = FALSE) # don't show multivariate tests
```








***
## Average Carbohydrate
### Linear mixed-effects model
```{r}
lme.model <- lme(AvgCHO ~ Week, 
                 random = ~1|Subject/Week, 
                 data=obs)
summary(lme.model)
```

```{r}
# print F- and p-values
anova(lme.model)
```

### Boxplot
```{r}
aboxplot <- ggplot(data=obs, aes(x=Week,y=AvgCHO)) + 
            geom_boxplot() +
            annotate(geom="text", label="*", x=2.01, y=190, size=6, family="serif") +
            labs(x="", y=expression("Mean  Carbohydrates g/day"))

aboxplot
```

### Pairwise t-test
```{r}
pairwise.t.test(x = obs$AvgCHO,
                g = obs$Week,
                p.adjust.method = "bonferroni",
                pool.sd = FALSE,
                paired = TRUE )
```
### Multivariate approach using lm() / mlm()
1. Convert data frame to matrix. Each row of the matrix is a subject.  Each column is the test number.
```{r}
obs.matrix <- with(obs, # data from obs data frame
                    cbind(
                      AvgCHO[Week=="Baseline"], # subject observations baseline
                      AvgCHO[Week=="Week1"], # subject observations test 1
                      AvgCHO[Week=="Week2"], # etc
                      AvgCHO[Week=="Week3"]
                      ) # cbind column bind, rows are flipped to columns
                  )
# set row and column identities
rownames(obs.matrix) <- levels(obs$Subject)
colnames(obs.matrix) <- levels(obs$Week)

# display the matrix
obs.matrix
```
2. Create a multivariate linear model from the matrix.
```{r}
mlm.model <- lm(obs.matrix ~ 1)
mlm.model

```

The coefficients are equal to the means across subjects at each test.

3. Create a set of factors.
```{r}
rfactor <- as.factor(levels(obs$Week))
rfactor
```

4. Define a new anova model object ```mlm.model.aov```.
```{r}
mlm.model.aov <- Anova(mlm.model, # the multivariate linear model
                       idata = data.frame(rfactor),
                       idesign = ~rfactor, 
                       type = "III")

summary(mlm.model.aov, 
        multivariate = FALSE) # don't show multivariate tests
```








***
## Average Protein
### Linear mixed-effects model
```{r}
lme.model <- lme(AvgPro ~ Week, 
                 random = ~1|Subject/Week, 
                 data=obs)
summary(lme.model)
```

```{r}
# print F- and p-values
anova(lme.model)
```

### Boxplot
```{r}
aboxplot <- ggplot(data=obs, aes(x=Week,y=AvgPro)) + 
            geom_boxplot() +
            labs(x="", y=expression("Mean Protein g/day"))

aboxplot
```

### Pairwise t-test
```{r}
pairwise.t.test(x = obs$AvgPro,
                g = obs$Week,
                p.adjust.method = "bonferroni",
                pool.sd = FALSE,
                paired = TRUE )
```
### Multivariate approach using lm() / mlm()
1. Convert data frame to matrix. Each row of the matrix is a subject.  Each column is the test number.
```{r}
obs.matrix <- with(obs, # data from obs data frame
                    cbind(
                      AvgPro[Week=="Baseline"], # subject observations baseline
                      AvgPro[Week=="Week1"], # subject observations test 1
                      AvgPro[Week=="Week2"], # etc
                      AvgPro[Week=="Week3"]
                      ) # cbind column bind, rows are flipped to columns
                  )
# set row and column identities
rownames(obs.matrix) <- levels(obs$Subject)
colnames(obs.matrix) <- levels(obs$Week)

# display the matrix
obs.matrix
```
2. Create a multivariate linear model from the matrix.
```{r}
mlm.model <- lm(obs.matrix ~ 1)
mlm.model

```

The coefficients are equal to the means across subjects at each test.

3. Create a set of factors.
```{r}
rfactor <- as.factor(levels(obs$Week))
rfactor
```

4. Define a new anova model object ```mlm.model.aov```.
```{r}
mlm.model.aov <- Anova(mlm.model, # the multivariate linear model
                       idata = data.frame(rfactor),
                       idesign = ~rfactor, 
                       type = "III")

summary(mlm.model.aov, 
        multivariate = FALSE) # don't show multivariate tests
```
