---
title: "Repeated Measures Analysis of Variance"
output: html_notebook
---
# Repeated measures analysis of variance for low carb high fat diet
The methodology comes from https://gribblelab.wordpress.com/2009/03/09/repeated-measures-anova-using-r/


## Load the packages that will be used
Load the various packages.

```{r setup}

# packages for analysis of repeated measures
library(nlme)
library(car)

# document definition and construction
library(knitr)
library(rmarkdown)

#reading and writing files of various types (in this case, csv and text files)
library(readr)

# data manipulation
library(tidyr)
library(dplyr)
library(tibble)

# pipelining commands
library(magrittr)

```
## Prepare the data
Read the data from the CSV file (which was copied and pasted from the source Excel spreadsheet).  Sort the data by subject so things will line up later on.  Remove the unneeded columns (not critical to the process).
```{r}
obs <- read_csv("mep_observations.csv")

#sort data to keep the factors ordered for a later matrix
obs <- obs[order(obs$Subject),]

# these models work with factors
obs$Subject <- as.factor(obs$Subject)
obs$TestNumber <- as.factor(obs$TestNumber)
obs <- as_tibble(obs)

obs
```






## Analysis of variance MEP HR
- The dependent variable is ```MEPHR```.
- The test subject is ```Subject```.
- The factor is ```TestNumber```.

### Univariate with aov()
```{r}
aov.model <- aov(MEPHR ~ TestNumber + Error(Subject/TestNumber), data=obs)
summary(aov.model)
```
### Univariate approach using lme()
This method fits a linear mixed-effects model.
```{r}
lme.model <- lme(MEPHR ~ TestNumber, random = ~1|Subject/TestNumber, data=obs)
summary(lme.model)
```

Display F- and p-values for the factor effect.
```{r}
anova(lme.model)
```

### Multivariate approach using lm() / mlm()
1. Convert data frame to matrix. Each row of the matrix is a subject.  Each column is the test number.
```{r}
obs.matrix <- with(obs, # data from obs data frame
                    cbind(
                      MEPHR[TestNumber=="0"], # subject observations baseline
                      MEPHR[TestNumber=="1"], # subject observations test 1
                      MEPHR[TestNumber=="2"], # etc
                      MEPHR[TestNumber=="3"]
                      ) # cbind column bind, rows are flipped to columns
                  )
# set row and column identities
rownames(obs.matrix) <- levels(obs$Subject)
colnames(obs.matrix) <- levels(obs$TestNumber)

# display the matrix
obs.matrix
```
2. Create a multivariate linear model from the matrix.
```{r}
mlm.model <- lm(obs.matrix ~ 1)
mlm.model

```

The coefficients are equal to the means across subjects at each test.

3. Create a set of factors.
```{r}
rfactor <- as.factor(levels(obs$TestNumber))
rfactor
```

4. Define a new anova model object ```mlm.model.aov```.
```{r}
mlm.model.aov <- Anova(mlm.model, # the multivariate linear model
                       idata = data.frame(rfactor),
                       idesign = ~rfactor, 
                       type = "III")

summary(mlm.model.aov, 
        multivariate = FALSE) # don't show multivariate tests
```







## Analysis of variance Weight
- The dependent variable is ```Weight```.
- The test subject is ```Subject```.
- The factor is ```TestNumber```.

### Univariate with aov()
```{r}
aov.model <- aov(Weight ~ TestNumber + Error(Subject/TestNumber), data=obs)
summary(aov.model)
```
### Univariate approach using lme()
This method fits a linear mixed-effects model.
```{r}
lme.model <- lme(Weight ~ TestNumber, random = ~1|Subject/TestNumber, data=obs)
summary(lme.model)
```

Display F- and p-values for the factor effect.
```{r}
anova(lme.model)
```

### Multivariate approach using lm() / mlm()
1. Convert data frame to matrix. Each row of the matrix is a subject.  Each column is the test number.
```{r}
obs.matrix <- with(obs, # data from obs data frame
                    cbind(
                      Weight[TestNumber=="0"], # subject observations baseline
                      Weight[TestNumber=="1"], # subject observations test 1
                      Weight[TestNumber=="2"], # etc
                      Weight[TestNumber=="3"]
                      ) # cbind column bind, rows are flipped to columns
                  )
# set row and column identities
rownames(obs.matrix) <- levels(obs$Subject)
colnames(obs.matrix) <- levels(obs$TestNumber)

# display the matrix
obs.matrix
```
2. Create a multivariate linear model from the matrix.
```{r}
mlm.model <- lm(obs.matrix ~ 1)
mlm.model

```

The coefficients are equal to the means across subjects at each test.

3. Create a set of factors.
```{r}
rfactor <- as.factor(levels(obs$TestNumber))
rfactor
```

4. Define a new anova model object ```mlm.model.aov```.
```{r}
mlm.model.aov <- Anova(mlm.model, # the multivariate linear model
                       idata = data.frame(rfactor),
                       idesign = ~rfactor, 
                       type = "III")

summary(mlm.model.aov, 
        multivariate = FALSE) # don't show multivariate tests
```








## Analysis of variance MaxRelVO2
- The dependent variable is ```MaxRelVO2```.
- The test subject is ```Subject```.
- The factor is ```TestNumber```.

### Univariate with aov()
```{r}
aov.model <- aov(MaxRelVO2 ~ TestNumber + Error(Subject/TestNumber), data=obs)
summary(aov.model)
```
### Univariate approach using lme()
This method fits a linear mixed-effects model.
```{r}
lme.model <- lme(MaxRelVO2 ~ TestNumber, random = ~1|Subject/TestNumber, data=obs)
summary(lme.model)
```

Display F- and p-values for the factor effect.
```{r}
anova(lme.model)
```

### Multivariate approach using lm() / mlm()
1. Convert data frame to matrix. Each row of the matrix is a subject.  Each column is the test number.
```{r}
obs.matrix <- with(obs, # data from obs data frame
                    cbind(
                      MaxRelVO2[TestNumber=="0"], # subject observations baseline
                      MaxRelVO2[TestNumber=="1"], # subject observations test 1
                      MaxRelVO2[TestNumber=="2"], # etc
                      MaxRelVO2[TestNumber=="3"]
                      ) # cbind column bind, rows are flipped to columns
                  )
# set row and column identities
rownames(obs.matrix) <- levels(obs$Subject)
colnames(obs.matrix) <- levels(obs$TestNumber)

# display the matrix
obs.matrix
```
2. Create a multivariate linear model from the matrix.
```{r}
mlm.model <- lm(obs.matrix ~ 1)
mlm.model

```

The coefficients are equal to the means across subjects at each test.

3. Create a set of factors.
```{r}
rfactor <- as.factor(levels(obs$TestNumber))
rfactor
```

4. Define a new anova model object ```mlm.model.aov```.
```{r}
mlm.model.aov <- Anova(mlm.model, # the multivariate linear model
                       idata = data.frame(rfactor),
                       idesign = ~rfactor, 
                       type = "III")

summary(mlm.model.aov, 
        multivariate = FALSE) # don't show multivariate tests
```









## Analysis of variance FatBIA
- The dependent variable is ```FatBIA```.
- The test subject is ```Subject```.
- The factor is ```TestNumber```.

### Univariate with aov()
```{r}
aov.model <- aov(FatBIA ~ TestNumber + Error(Subject/TestNumber), data=obs)
summary(aov.model)
```
### Univariate approach using lme()
This method fits a linear mixed-effects model.
```{r}
lme.model <- lme(FatBIA ~ TestNumber, random = ~1|Subject/TestNumber, data=obs)
summary(lme.model)
```

Display F- and p-values for the factor effect.
```{r}
anova(lme.model)
```

### Multivariate approach using lm() / mlm()
1. Convert data frame to matrix. Each row of the matrix is a subject.  Each column is the test number.
```{r}
obs.matrix <- with(obs, # data from obs data frame
                    cbind(
                      FatBIA[TestNumber=="0"], # subject observations baseline
                      FatBIA[TestNumber=="1"], # subject observations test 1
                      FatBIA[TestNumber=="2"], # etc
                      FatBIA[TestNumber=="3"]
                      ) # cbind column bind, rows are flipped to columns
                  )
# set row and column identities
rownames(obs.matrix) <- levels(obs$Subject)
colnames(obs.matrix) <- levels(obs$TestNumber)

# display the matrix
obs.matrix
```
2. Create a multivariate linear model from the matrix.
```{r}
mlm.model <- lm(obs.matrix ~ 1)
mlm.model

```

The coefficients are equal to the means across subjects at each test.

3. Create a set of factors.
```{r}
rfactor <- as.factor(levels(obs$TestNumber))
rfactor
```

4. Define a new anova model object ```mlm.model.aov```.
```{r}
mlm.model.aov <- Anova(mlm.model, # the multivariate linear model
                       idata = data.frame(rfactor),
                       idesign = ~rfactor, 
                       type = "III")

summary(mlm.model.aov, 
        multivariate = FALSE) # don't show multivariate tests
```










## Analysis of variance FatSF
- The dependent variable is ```FatSF```.
- The test subject is ```Subject```.
- The factor is ```TestNumber```.

### Univariate with aov()
```{r}
aov.model <- aov(FatSF ~ TestNumber + Error(Subject/TestNumber), data=obs)
summary(aov.model)
```
### Univariate approach using lme()
This method fits a linear mixed-effects model.
```{r}
lme.model <- lme(FatSF ~ TestNumber, random = ~1|Subject/TestNumber, data=obs)
summary(lme.model)
```

Display F- and p-values for the factor effect.
```{r}
anova(lme.model)
```

### Multivariate approach using lm() / mlm()
1. Convert data frame to matrix. Each row of the matrix is a subject.  Each column is the test number.
```{r}
obs.matrix <- with(obs, # data from obs data frame
                    cbind(
                      FatSF[TestNumber=="0"], # subject observations baseline
                      FatSF[TestNumber=="1"], # subject observations test 1
                      FatSF[TestNumber=="2"], # etc
                      FatSF[TestNumber=="3"]
                      ) # cbind column bind, rows are flipped to columns
                  )
# set row and column identities
rownames(obs.matrix) <- levels(obs$Subject)
colnames(obs.matrix) <- levels(obs$TestNumber)

# display the matrix
obs.matrix
```
2. Create a multivariate linear model from the matrix.
```{r}
mlm.model <- lm(obs.matrix ~ 1)
mlm.model

```

The coefficients are equal to the means across subjects at each test.

3. Create a set of factors.
```{r}
rfactor <- as.factor(levels(obs$TestNumber))
rfactor
```

4. Define a new anova model object ```mlm.model.aov```.
```{r}
mlm.model.aov <- Anova(mlm.model, # the multivariate linear model
                       idata = data.frame(rfactor),
                       idesign = ~rfactor, 
                       type = "III")

summary(mlm.model.aov, 
        multivariate = FALSE) # don't show multivariate tests
```

